A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods

### 2024-06-18 Draft/Draft_MEM.tex
This is the re-submitted paper.x

\title{A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods}
\author{Johannes Bracher$^{1, 2}$, Jonas M. Littek$^1$}
\date{9 June 2023\\ \medskip \small
$^1$Karlsruhe Institute of Technology (KIT), Institute of Statistics\\
$^2$Heidelberg Institute for Theoretical Studies (HITS), Computational Statistics Group\\ \medskip
Email address for correspondence:\\ \url{johannes.bracher@kit.edu}\\ % \medskip
}

@@

\abstract{
\begin{quote}
Monitoring of influenza activity is a key task of public health agencies around the world. Intensity thresholds are commonly used to retrospectively classify season peak intensity and to compare current influenza activity to the peak intensity of past seasons. The resulting classifications into low, medium, high or very high intensity help to inform national risk assessment, public health planning and ressource allocation. Moreover, they feed into global risk analyses like the weekly World Health Organization (WHO) Influenza Update. Two common intensity thresholding approaches, recommended in dedicated WHO guidelines, are the moving epidemic method (MEM) and the WHO method. In both methods, thresholds correspond to quantiles of a normal distribution which has been fitted to a set of potentially transformed historical observations. While an extensive body of literature on practical applications of the MEM and WHO method exists, their statistical properties have not been assessed systematically. In this paper we study the two methods analytically and in a simulation study based on re-sampling of data from French and US routine influenza surveillance. Moreover, straightforward extensions to account for small sample sizes and secular trends are described. We find that under the default settings, both the MEM and WHO method on average classify more season peaks than intended as high or very high intensity. Combining characteristics of both and adding a standard correction for small sample sizes, better-calibrated thresholds can be achieved. Even these, however, have modest sensitivity and positive predictive values. This concerns especially thresholds for the very high intensity level, which should thus be interpreted with care.
\end{quote}
}

@@

\begin{quote}
\textbf{Keywords:} calibration, influenza, intensity threshold, moving epidemic method, predictive quantile, re-sampling, sensitivity, WHO method.
\end{quote}

@@

Influenza is a major public health threat and monitoring of influenza activity is a central task of health authorities around the world. To strengthen and standardize monitoring in the aftermath of the 2009 influenza H1N1 pandemic, the World Health Organisation (WHO) developed recommendations on severity assessment during seasonal and pandemic outbreaks. According to the PISA guidelines (\textit{Pandemic Influenza Severity Assessment}, \citealt{WHO2014}), severity is defined in terms of three dimensions. \textit{Transmissibility} refers to how many people become ill and is measured e.g., via weekly case numbers of influenza-like illness (ILI) or the percentage of such cases among all consultations by general practitioners. To measure \textit{severity}, ratios of recorded hospitalizations and cases or deaths and hospitalizations can be used. Lastly, \textit{impact} is measured e.g., by weekly case numbers of severe acute respiratory disease (SARI) or admissions to intensive care units.

A key role in the WHO PISA guideline is taken by influenza intensity thresholds. For transmissibility and impact indicators, they represent a generic tool to classify the peak intensity of an influenza season in light of past peak values. Typically, classifications into \textit{low}, \textit{medium/moderate}, \textit{high} and \textit{very high/extraordinary} intensity are provided. The rationale is that
\begin{quote}
\textit{``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold''} \citep[p.10]{WHO2017}.
\end{quote}
Similarly to the idea of a ``100-year-flood'' used to communicate hydrological risks \citep{Holmes2010}, the definition of high and very high intensity influenza seasons thus implies that on average they occur every 10 and 40 years, respectively.

Intensity thresholds form part of monitoring procedures in numerous countries, contributing to national-level risk appraisal and planning; see e.g., \cite{CDC2024} on their use in the United States. In addition, intensity classifications feed into international situation assessments like the weekly \textit{WHO Influenza Update} (see e.g., \citealt{WHO2024}). In this context, they facilitate comparisons and summaries across countries and surveillance systems. In practice, two uses of thresholds can be distinguished \citep{CDC2024}:
\begin{itemize}
\item[(1)] In the aftermath of an influenza season, thresholds serve to classify its peak intensity. As an example, the top panel of Figure \ref{fig:maps} shows an assessment of the 2022/23 season by the US Centers for Disease Prevention and Control (CDC; \citealt{White2023}). In terms of influenza-associated outpatient visits, hospitalization rates, and deaths among children and adolescents, this season was judged as high intensity.
\item[(2)] During an influenza season, values of relevant indicators can be compared to intensity thresholds on a weekly basis. The resulting classification is relative to past peaks rather than typical values at a given time of the season. This use of thresholds is illustrated in the bottom panel of Figure \ref{fig:maps}. This display, published by WHO and the European Centre for Disease Prevention and Control (ECDC) summarizes influenza activity across Europe in a map and a heat chart.
\end{itemize}

\begin{figure}[h!]
\begin{center}
\fbox{\includegraphics[width=0.65\textwidth]{figure/mmwr_with_heading.png}}

@@

\fbox{
\includegraphics[width=0.645\textwidth]{figure/map_plus_table.png}}
\end{center}
\caption{Use of influenza intensity thresholds by public health agencies. Top: Retrospective classifications for the 2022/23 influenza season in the US by the CDC \citep{White2023}. Here, $\text{IT}_{50}, \text{IT}_{90}$ and $\text{IT}_{98}$ denote thresholds for medium, high and very high intensity. Bottom: Influenza intensity map and heat chart as published by WHO and ECDC in early 2023 (\url{https://flunewseurope.org/}). The display is collated from multiple figures and the heat chart is cropped from a list of all countries in the European Region of WHO. Copyright information is provided at the end of this article.
}
\label{fig:maps}
\end{figure}

@@

In the PISA guidelines \citep{WHO2017}, two statistical methods were recommended for influenza intensity thresholding. These are the \textit{moving epidemic method} (MEM; \citealt{Vega2015}) and the \textit{WHO method} \citep{WHO2014}, also referred to as the \textit{average curves method}. The latter has also been recommended by ECDC (e.g., \citealt{ECDC2017}) and is used by the US CDC \citep{Biggerstaff2017}. As will be detailed in Section \ref{sec:definitions}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach. The two methods have been adopted by many public health agencies (see Section \ref{sec:review}). Their statistical properties, however, have not yet been studied. In the present work, we aim to close this gap and derive some recommendations for the computation of influenza intensity thresholds. We study the MEM and WHO methods and suggest extensions to account for small sample sizes and secular trends. We obtain some analytical results on their properties, complemented by simulation experiments based on re-sampling of routine surveillance data from France and the US. These indicate that using the default settings, both the MEM and the WHO method tend to produce too low thresholds. Consequently, the number of season peaks classified as high or very high intensity is higher than intended. The best-behaved thresholds are achieved by combining characteristics of the MEM and WHO methods, along with a correction for small sample sizes. Even this configuration, however, is characterized by modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns especially thresholds for very high intensity, which should thus be interpreted with care.

@@

\begin{description}
\item \textbf{Calibration.} As stated in Section \ref{sec:introduction}, thresholds are defined in a statistical manner -- the respective thresholds should be exceeded by 60\%, 10\% and 2.5\% of season peaks. Threshold setting thus corresponds to determining predictive quantiles for future season peak values, and a major requirement is that these are \textit{calibrated}. We call a thresholding method calibrated if in the long run, the intended fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the intended fraction of seasons into the various categories, the individual classifications should be correct. Assuming there is an underlying stationary distribution of season peaks, a peak which is truly among the 2.5\%/10\%/60\% highest peaks should also be flagged as such. Conversely, a peak which is not actually among the 2.5\%/10\%/60\% highest ones should not exceed the respective threshold. This aspect, though, can only be assessed in theory and simulation studies, where the true intensity status can be determined based on the assumed distribution of season peaks. In real-world applications, there is no ``gold standard'' to determine the true intensity level.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method and its parameters should not require advanced statistical training.
\item \textbf{Ease of practical application.} Methods should be straightforward to apply using well-documented and, ideally, open-source software packages.
\end{description}

@@

\noindent The MEM and WHO methods are special cases of this procedure, see also Table \ref{tab:differences} and the graphical illustration in Figure \ref{fig:illustration}. In the MEM, no smoothing is applied and the default transformation $f$ is the natural logarithm.  \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons to ensure a recent data basis. The number of observations included per season is set to $n = 30/m$, rounded to the nearest integer (with a minimum value of $n = 1$). The total number $m \times n$ of historical observations is thus kept approximately fixed. This corresponds to the description by \cite{Vega2015} and the default settings of the R package \texttt{mem} \citep{Lozano2020}. The package, however, permits the user to choose $n$, $m$, and $f$, see Supplement B.1. We note that while \texttt{mem} does not currently allow for data smoothing using a moving average, some alternatives are available. Thresholds based on a $t$-distribution as in equation \eqref{eq:q_Y_t} were not implemented when the computations for this paper were run, and an independent custom function was used instead. In the meantime, however, this feature has been added to the development version of the \texttt{mem} package, see Supplement B.1.

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.78\textwidth]{figure/illustration_mem_who.pdf}
\end{center}
\caption{Illustration of the MEM and WHO methods with their respective default settings, computed using $m = 6$ historical seasons (solid lines) and applied to a seventh (dotted lines). Here, both methods classify the new season peak as medium intensity. Observations included in the reference set are highlighted by dots (with $n = 30/6 = 5$ for MEM, $n = 1$ smoothed values for WHO). The distribution fitted to the reference set is shown as a vertical density plot. Note that for the MEM this is a log-normal as the normal distribution is fitted to log-transformed values. The data used for illustration correspond to the years 2012--2019 from the French region of Grand Est, see Section \ref{subsec:data} and Figure \ref{fig:data}.}
\label{fig:illustration}
\end{figure}

@@

The EARS approach can also be seen as an intercept-only linear regression model,
$$
Y_{t} = \beta_0 + \varepsilon_{t}, \ \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
$$
with $\bar{y}$ and $s^2$ as estimators of $\beta_0$ and $\sigma_\varepsilon^2$. Indeed, regression-based approaches represent a flexible and general monitoring framework, compare Section 2 in \cite{Unkel2012}. Classic parametric approaches like the \textit{Serfling method} \citep{Serfling1896} are widely used, see e.g., \cite{Thompson2009} for an application to influenza mortality. Here, a regression model including time trends and sinusoidal functions for seasonality, say,
\begin{equation}
Y_{t} = \beta_0 + \beta_1 t + \beta_2 \sin(2\pi t / \omega) + \beta_3 \cos(2\pi t / \omega) + \varepsilon_t, \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
\label{eq:serfling}
\end{equation}
is fitted to historical values $y_1, \dots, y_m$ spanning multiple seasons. The period of seasonal cycles is given by $\omega$, with $\omega = 52$ for weekly data and yearly seasonality. From the model fit predictive quantiles for a new value $y_{m + 1}$ are obtained, which serve as intensity thresholds. The equally popular \textit{Farrington method} is conceptually similar, but adapts to low-count settings via suitable data transformations \citep{Farrington1996} or count-valued response distributions \citep{Noufaily2013}. While technically a generalization of the EARS approach, the rationale of the Serfling and Farrington methods is somewhat different. As noted by \cite{Hutwagner2003}, the latter aim to detect deviations from past seasonal patterns while accounting for long-term trends. EARS, on the other hand, targets abrupt changes relative to the most recent observations.

@@

We now re-state the purpose and assumptions of the WHO and moving epidemic methods in statistical terms, denoting random variables by capital letters (e.g., $X_j^{(i)}$) and their realizations by lower-case letters (e.g., $x_j^{(i)}$). The goal of the thresholding exercise is to use data from seasons 1 through $m$ to obtain values $\hat{q}_{Y, \alpha}$ which shall be exceeded by a new transformed peak value $Y_{m + 1}^{(1)}$ with probability $(1 - \alpha) \in \{0.6, 0.1, 0.025\}$, respectively. The $\hat{q}_{Y, \alpha}$ can thus be seen as \textit{predictive quantiles} at levels $\alpha \in \{0.4, 0.9, 0.975\}$. Prediction of new realizations based on parametric assumptions and a set of previous observations is a well-studied problem; see e.g., \cite{Millard2013} on applications in environmental studies. We note that in \cite{WHO2014} and \cite{Vega2015} the thresholds are refered to as the upper ends of one-sided confidence intervals. This, however, is imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ is used (see documentation of the \texttt{mem} package and \citealt{WHO2014}, p.69).

@@

As mentioned in Secion \ref{sec:related_literature}, regression approaches are a natural extension of moment-based techniques like the WHO and moving epidemic method. We illustrate this with a suggestion to account for secular trends in threshold setting, thus relaxing assumption (HB). Secular trends are the reason why for the MEM it is recommend to use no more than $m = 10$ historical seasons, with \citeauthor{Vega2013} (\citeyear{Vega2013}, p.556) cautioning that ``more than ten seasons may further increase accuracy but make a model susceptible to biases from secular trends.'' An obvious model accouning for a trend is
\begin{equation}
Y_j^{(1)} = \beta_0 + \beta_1 \times j + \varepsilon_j, \ \ \ \ \varepsilon_j \stackrel{\text{i.i.d.}}{\sim} \text{N}(0, \sigma^2_\varepsilon).\label{eq:trend}
\end{equation}
Parameter estimates $\hat{\alpha}$ and $\hat{\beta}$ can be computed using peak values $y_1^{(1)}, \dots, y_m^{(1)}$ from the last $m$ seasons and the ordinary least squares method. An unbiased estimator of ${\sigma}^2_\varepsilon$ is
$$
\hat{\sigma}^2_\varepsilon = \frac{1}{m - 2} \sum_{j = 1}^m (y^{(1)}_i - \beta_0 - \beta_1 \times j)^2.
$$
Predictive quantiles at levels $\alpha = 0.4, 0.9, 0.975$, and thus intensity thresholds, can be obtained using standard methods implemented in software packages such as R. As we show in Supplement B.2, the generalization of equation \eqref{eq:q_Y_t} resulting for model \eqref{eq:trend} is of a simple form. A graphical illustration of the resulting time-varying thresholds can be found in Figure \ref{fig:illustration_trend}. This approach could easily be extended to accommodate $n > 1$ values per historical season. However, as we will argue in Section \ref{subsec:choice_n}, this will lead to biased thresholds, unless a more flexible model formulation accounting for relevant mean and correlation structures is adopted. We therefore only consider $n = 1$.

@@

An obvious question is under which conditions thresholds should be corrected for secular trends. In\cite{Farrington1996}, a trend is included if at least three years of weekly data are available and the trend parameter is significant at the 5\% level. This, however, does not seem practical in intensity thresholding, where historical peak values are scarce. The related question of how many observations are needed to reliably fit model \eqref{eq:trend} has been discussed by \cite{Hyndman2007}. However, the authors conclude that this depends on $\beta_1/\sigma_\varepsilon$, meaning that there cannot be any general recommendation. We will return to this question in Section \ref{subsec:results_trends} and the Discussion.

@@

\caption{Sensitivity, specificity and positive predictive values under normality assumption, by number $m$ of historical seasons used. Top row: when using the original formulation \eqref{eq:def_q} based on the normal distribution. Bottom: when using formulation \eqref{eq:q_Y_t} based on the $t$-distribution. Dashed lines show the approximations \eqref{eq:sens}--\eqref{eq:ppv}, solid lines simulation results.}
\label{fig:sens_spec_ana}
\end{figure}

@@

We use publicly available data on the estimated weekly incidence of influenza-like illness per 100,000 inhabitants in France, 1986--2019, published by Réseau Sentinelles (INSERM/Sorbonne Université, \url{https://www.sentiweb.fr}, \citealt{Flahault2006}). These are available at the national and regional levels. To make statements about the ``very high'' categorization, we require a larger set of historical data than the available 34 seasons. We therefore pool curves from the 12 continental French administrative regions. As the overall level of ILI incidences varies considerably across regions, we scale all data such that the average season peak per region is 100. As the incidence curves from the Corse island region differ substantially from the other regions we exclude them. Moreover, the pandemic 2009/2010 season was removed. In total we then dispose of $12 \times 33 = 396$ season curves.

@@

\textbf{Confusion matrices.} To complement these results and provide some more intuition on the intensity classifications, Figure \ref{fig:mosaic} shows confusion matrices for thresholds computed with a log transformation and different values of $m$. This represents a more detailed breakup of the results from the second column of Figure \ref{fig:results1}. Note that the perspective differs somewhat from the computation of sensitivities and specificities. There, we focused on \textit{threshold exceedance}, considering for instance whether the medium threshold was exceeded by a peak which was truly at least medium intensity. For the sensitivity of the medium threshold we thus ignored whether a peak which was truly medium intensity also exceeded the high threshold (which was in turn reflected in the specificity of the high threshold). Now we consider not just threshold exceedance, but the exact categorization. This is shown stratified by the true category of a peak, determined via the empirical distribution of all 396 available peaks (the highest 2.5\% are considered very high, the next 7.5\% high, the next 50\% medium and the remaining 40\% low).

>>>>

\title{A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods}
\author{Johannes Bracher$^{1, 2}$, Jonas M. Littek$^1$}
\date{18 June 2024\\ \medskip \small
$^1$Karlsruhe Institute of Technology (KIT), Institute of Statistics\\
$^2$Heidelberg Institute for Theoretical Studies (HITS), Computational Statistics Group\\ \medskip
Email address for correspondence:\\ \url{johannes.bracher@kit.edu}\\ % \medskip
}

@@

\abstract{
\begin{quote}
Monitoring of influenza activity is a key task of public health agencies around the world. Intensity thresholds serve to retrospectively classify season peak intensity and to compare current influenza activity to past peak values. The resulting classifications into low, medium, high or very high intensity inform national-level risk assessment and planning. Moreover, they feed into international summary reports. Two common thresholding approaches, recommended in dedicated WHO guidelines, are the moving epidemic method (MEM) and the WHO method. In both approaches, thresholds correspond to quantiles of a normal distribution fitted to a set of historical observations. While an extensive literature on applications of these methods exists, their statistical properties have not been assessed systematically. In this paper we study them analytically and in a simulation study based on re-sampling of data from French influenza surveillance. Moreover, extensions to account for small sample sizes and secular trends are described. Under the default settings, both the MEM and WHO method on average classify more seasons than intended as high or very high intensity. Combining characteristics of both and adding a small-sample correction, better-calibrated thresholds can be achieved. Even these, however, have modest sensitivity and positive predictive values. This concerns especially thresholds for very high intensity.
\end{quote}
}

@@

\begin{quote}
\textbf{Keywords:} calibration, influenza intensity threshold, moving epidemic method, predictive quantile, re-sampling, WHO method.
\end{quote}

@@

Influenza is a major public health threat and monitoring of influenza activity is a central task of health authorities around the world. To strengthen and standardize monitoring in the aftermath of the 2009 influenza H1N1 pandemic, the World Health Organisation (WHO) developed recommendations on severity assessment during seasonal and pandemic outbreaks. According to the PISA guidelines (\textit{Pandemic Influenza Severity Assessment}, \citealt{WHO2014}), severity is defined in terms of three dimensions. \textit{Transmissibility} refers to how many people become ill and is measured e.g., via weekly case numbers of influenza-like illness (ILI) or the percentage of such cases among all consultations by general practitioners. \textit{Severity} is commonly assessed by ratios of recorded deaths and hospitalizations or hospitalizations and cases. Lastly, \textit{impact} is measured e.g., by weekly case numbers of severe acute respiratory disease (SARI) or admissions to intensive care units.

A key role in the WHO PISA guideline is taken by influenza intensity thresholds. For transmissibility and impact indicators, they represent a generic tool to classify the peak intensity of an influenza season in light of past peak values. Typically, classifications into \textit{low}, \textit{medium/moderate}, \textit{high} and \textit{very high/extraordinary} intensity are provided. The rationale is that
\begin{quote}
\textit{``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold''} \citep[p.10]{WHO2017}.
\end{quote}
Similarly to the idea of a ``100-year-flood'' used to communicate hydrological risks \citep{Holmes2010}, the definition of high and very high intensity influenza season peaks thus implies that on average they occur every 10 and 40 years, respectively.

Intensity thresholds form part of monitoring procedures in numerous countries, contributing to national-level risk appraisal and planning; see e.g., \cite{CDC2024} on their use in the United States. In addition, intensity classifications feed into international situation assessments like the weekly \textit{WHO Influenza Update} (see e.g., \citealt{WHO2024}). In this context, they facilitate comparisons and summaries across countries and surveillance systems. In practice, two uses of thresholds can be distinguished \citep{CDC2024}:
\begin{itemize}
\item[(1)] In the aftermath of an influenza season, thresholds serve to classify its peak intensity. As an example, the top panel of Figure \ref{fig:maps} shows an assessment of the 2022/23 season by the US Centers for Disease Prevention and Control (CDC; \citealt{White2023}). In terms of influenza-associated outpatient visits, hospitalization rates, and deaths among children and adolescents, this season was judged as high intensity.
\item[(2)] During an influenza season, values of relevant indicators can be compared to intensity thresholds on a weekly basis. The resulting classification is relative to past peaks rather than typical values at a given time of the season. An illustration is provided in the bottom panel of Figure \ref{fig:maps}. This display, published by the European Centre for Disease Prevention and Control (ECDC), summarizes influenza activity across Europe in a stylized map with colour coding.
\end{itemize}

\begin{figure}[h!]
\begin{center}
\fbox{\includegraphics[width=0.85\textwidth]{figure/mmwr_with_heading.png}}

@@

\fbox{
\includegraphics[width=0.845\textwidth]{figure/bubbles.png}}
\end{center}
\caption{Use of influenza intensity thresholds by public health agencies. Top: Retrospective classifications for the 2022/23 influenza season in the US by the CDC \citep{White2023}. Here, $\text{IT}_{50}, \text{IT}_{90}$ and $\text{IT}_{98}$ denote thresholds for medium, high and very high intensity. Bottom: Stylized influenza intensity maps published by ECDC in weeks 4 through 6, 2020. Copyright information is provided at the end of this article.
}
\label{fig:maps}
\end{figure}

@@

In the PISA guidelines \citep{WHO2017}, two statistical methods were recommended for influenza intensity thresholding. These are the \textit{moving epidemic method} (MEM; \citealt{Vega2015}) and the \textit{WHO method} \citep{WHO2014}, also referred to as the \textit{average curves method}. The former has also been recommended by ECDC (e.g., \citealt{ECDC2017}) and is used by the US CDC \citep{Biggerstaff2017}. As will be detailed in Section \ref{sec:definitions}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach. The two methods have been adopted by many public health agencies (see Section \ref{sec:review}). Their statistical properties, however, have not yet been studied. In the present work, we aim to close this gap and derive some recommendations for the computation of influenza intensity thresholds. We study the MEM and WHO methods and suggest extensions to account for small sample sizes and secular trends. We obtain some analytical results on their properties, complemented by simulation experiments based on re-sampling of routine surveillance data from France and the US. These indicate that using the default settings, both the MEM and the WHO method tend to produce too low thresholds. Consequently, the number of season peaks classified as high or very high intensity is higher than intended. The best-behaved thresholds are achieved by combining characteristics of the MEM and WHO methods, along with a correction for small sample sizes. Even this configuration, however, is characterized by modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns especially thresholds for very high intensity, which should thus be interpreted with care.

@@

\begin{description}
\item \textbf{Calibration.} As stated in Section \ref{sec:introduction}, thresholds are defined in a statistical manner -- the respective thresholds should be exceeded by 60\%, 10\% and 2.5\% of season peaks. Threshold setting thus corresponds to determining predictive quantiles for future season peak values, and a major requirement is that these are \textit{calibrated}. We call a thresholding method calibrated if in the long run, the intended fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the intended fraction of seasons into the various categories, the individual classifications should be correct. Assuming there is an underlying stationary distribution of season peaks, a peak which is truly among the 2.5\%/10\%/60\% highest peaks should also be flagged as such (\textit{sensitivity}). Conversely, a peak which is not actually among the 2.5\%/10\%/60\% highest ones should not exceed the respective threshold (\textit{specificity}). These aspects, though, can only be assessed in theory and simulation studies, where the true intensity status can be determined based on the assumed distribution of peaks. In real-world applications, there is no ``gold standard'' to determine the true intensity level.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method and its parameters should not require advanced statistical training.
\item \textbf{Ease of practical application.} Methods should be straightforward to apply using well-documented and, ideally, open-source software packages.
\end{description}

@@

\noindent The MEM and WHO methods are special cases of this procedure, see also Table \ref{tab:differences} and the graphical illustration in Figure \ref{fig:illustration}. In the MEM, no smoothing is applied and the default transformation $f$ is the natural logarithm.  \cite{Vega2015} recommend using $5 \leq m \leq 10$ seasons to ensure a recent data basis. The number of observations included per season is set to $n = 30/m$, rounded to the nearest integer (with a minimum value of $n = 1$). The total number $m \times n$ of historical observations is thus kept approximately fixed. This corresponds to the description by \cite{Vega2015} and the default settings of the R package \texttt{mem} \citep{Lozano2020}. The package, however, permits the user to choose $n$, $m$, and $f$, see Supplement B.1. We note that while \texttt{mem} does not currently allow for data smoothing using a moving average, some alternatives are available. Thresholds based on a $t$-distribution as in equation \eqref{eq:q_Y_t} were not implemented when the computations for this paper were run, and an independent custom function was used instead. In the meantime, however, this feature has been added to the development version of the \texttt{mem} package, see Supplement B.1.

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.78\textwidth]{figure/illustration_mem_who.pdf}
\end{center}
\caption{Illustration of the MEM and WHO methods with their respective default settings, computed using $m = 6$ historical seasons (solid / dashed lines depending on whether smoothing is applied). These thresholds are applied to a seventh season (dotted lines, peak highlighted by a cross). Here, both methods classify the new season peak as medium intensity. Observations included in the reference set are highlighted by dots (with $n = 30/6 = 5$ for MEM, $n = 1$ smoothed values for WHO). The distribution fitted to the reference set is shown as a vertical density plot. Note that for the MEM this is a log-normal as the normal distribution is fitted to log-transformed values. The data used for illustration correspond to the years 2012--2019 from the French region of Grand Est, see Section \ref{subsec:data} and Figure \ref{fig:data}.}
\label{fig:illustration}
\end{figure}

@@

The EARS approach can also be seen as an intercept-only linear regression model,
$$
Y_{t} = \beta_0 + \varepsilon_{t}, \ \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
$$
with $\bar{y}$ and $s^2$ as estimators of $\beta_0$ and $\sigma_\varepsilon^2$. Indeed, regression-based approaches represent a flexible and general monitoring framework, compare Section 2 in \cite{Unkel2012}. Classic parametric approaches like the \textit{Serfling method} \citep{Serfling1896} are widely used, see e.g., \cite{Thompson2009} for an application to influenza mortality. Here, a regression model including time trends and sinusoidal functions for seasonality, say,
\begin{equation}
Y_{t} = \beta_0 + \beta_1 t + \beta_2 \sin(2\pi t / \omega) + \beta_3 \cos(2\pi t / \omega) + \varepsilon_t, \ \ \ \varepsilon_t \sim \text{N}(0, \sigma_\varepsilon^2),
\label{eq:serfling}
\end{equation}
is fitted to historical values $y_1, \dots, y_m$ spanning multiple seasons. The period of seasonal cycles is given by $\omega$, with $\omega = 52$ for weekly data and yearly seasonality. From the model fit predictive quantiles for a new value $y_{m + 1}$ are obtained, which serve as alarm thresholds. The equally popular \textit{Farrington method} is conceptually similar, but adapts to low-count settings via suitable data transformations \citep{Farrington1996} or count-valued response distributions \citep{Noufaily2013}. While technically a generalization of the EARS approach, the rationale of the Serfling and Farrington methods is somewhat different. As noted by \cite{Hutwagner2003}, the latter aim to detect deviations from past seasonal patterns while accounting for long-term trends. EARS, on the other hand, targets abrupt changes relative to the most recent observations.

@@

We now re-state the purpose and assumptions of the WHO and moving epidemic methods in statistical terms, denoting random variables by capital letters (e.g., $X_j^{(i)}$) and their realizations by lower-case letters (e.g., $x_j^{(i)}$). The goal of the thresholding exercise is to use data from seasons 1 through $m$ to obtain values $\hat{q}_{Y, \alpha}$ which shall be exceeded by a new transformed peak value $Y_{m + 1}^{(1)}$ with probabilities $(1 - \alpha) \in \{0.6, 0.1, 0.025\}$, respectively. The $\hat{q}_{Y, \alpha}$ can thus be seen as \textit{predictive quantiles} at levels $\alpha \in \{0.4, 0.9, 0.975\}$. Prediction of new realizations based on parametric assumptions and a set of previous observations is a well-studied problem; see e.g., \cite{Millard2013} on applications in environmental studies. We note that in \cite{WHO2014} and \cite{Vega2015} the thresholds are refered to as the upper ends of one-sided confidence intervals. This, however, is imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ is used (see documentation of the \texttt{mem} package and \citealt{WHO2014}, p.69).

@@

As mentioned in Secion \ref{sec:related_literature}, regression approaches are a natural extension of moment-based techniques like the WHO and moving epidemic methods. We illustrate this with a suggestion to account for secular trends in threshold setting, thus relaxing assumption (HB). Secular trends are the reason why for the MEM it is recommend to use no more than $m = 10$ historical seasons, with \citeauthor{Vega2013} (\citeyear{Vega2013}, p.556) cautioning that ``more than ten seasons may further increase accuracy but make a model susceptible to biases from secular trends.'' An obvious model accouning for a trend is
\begin{equation}
Y_j^{(1)} = \beta_0 + \beta_1 \times j + \varepsilon_j, \ \ \ \ \varepsilon_j \stackrel{\text{i.i.d.}}{\sim} \text{N}(0, \sigma^2_\varepsilon).\label{eq:trend}
\end{equation}
Parameter estimates $\hat{\alpha}$ and $\hat{\beta}$ can be computed using peak values $y_1^{(1)}, \dots, y_m^{(1)}$ from the last $m$ seasons and the ordinary least squares method. An unbiased estimator of ${\sigma}^2_\varepsilon$ is
$$
\hat{\sigma}^2_\varepsilon = \frac{1}{m - 2} \sum_{j = 1}^m (y^{(1)}_i - \beta_0 - \beta_1 \times j)^2.
$$
Predictive quantiles at levels $\alpha = 0.4, 0.9, 0.975$, and thus intensity thresholds, can be obtained using standard methods implemented in software packages such as R. As we show in Supplement B.2, the generalization of equation \eqref{eq:q_Y_t} resulting for model \eqref{eq:trend} is of a simple form. A graphical illustration of the resulting time-varying thresholds can be found in Figure \ref{fig:illustration_trend}. This approach could easily be extended to accommodate $n > 1$ values per historical season. However, as we will argue in Section \ref{subsec:choice_n}, this will lead to biased thresholds, unless a more flexible model formulation accounting for relevant mean and correlation structures is adopted. We therefore only consider $n = 1$.

@@

An obvious question is under which conditions thresholds should be corrected for secular trends. In \cite{Farrington1996}, a trend is included if at least three years of weekly data are available and the trend parameter is significant at the 5\% level. This, however, does not seem practical in intensity thresholding, where historical peak values are scarce. The related question of how many observations are needed to reliably fit model \eqref{eq:trend} has been discussed by \cite{Hyndman2007}. However, the authors conclude that this depends on $\beta_1/\sigma_\varepsilon$, meaning that there cannot be any general recommendation. We will return to this question in Section \ref{subsec:results_trends} and the Discussion.

@@

\caption{Sensitivity, specificity and PPVs with $n = 1$ and under the normality assumption (N), by number $m$ of historical seasons used. Top row: when using the original formulation \eqref{eq:def_q} based on the normal distribution. Bottom: when using formulation \eqref{eq:q_Y_t} based on the $t$-distribution. Dashed lines show the approximations \eqref{eq:sens}--\eqref{eq:ppv}, solid lines simulation results.}
\label{fig:sens_spec_ana}
\end{figure}

@@

We use publicly available data on the estimated weekly incidence of influenza-like illness per 100,000 inhabitants in France, 1986--2019, published by Réseau Sentinelles (INSERM/Sorbonne Université, \url{https://www.sentiweb.fr}, \citealt{Flahault2006}). These are available at the national and regional levels. To make statements about the ``very high'' category, we require a larger set of historical data than the available 34 seasons. We therefore pool curves from the 12 continental French administrative regions. As the overall level of ILI incidence varies considerably across regions, we scale all data such that the average season peak per region is 100. As the incidence curves from the Corse island region differ substantially from the other regions we exclude them. Moreover, the pandemic 2009/2010 season was removed. In total we then dispose of $12 \times 33 = 396$ season curves.

@@

\textbf{Confusion matrices.} To complement these results and provide some more intuition on the intensity classifications, Figure \ref{fig:mosaic} shows confusion matrices for thresholds computed with a log transformation and different values of $m$. This represents a more detailed breakup of the results from the second column of Figure \ref{fig:results1}. Note that the perspective differs somewhat from the computation of sensitivities and specificities. There, we focused on \textit{threshold exceedance}, considering for instance whether the medium threshold was exceeded by a peak which was truly at least medium intensity. For the sensitivity of the medium threshold we thus ignored whether a peak which was truly medium intensity also exceeded the high threshold (which was in turn reflected in the specificity of the high threshold). Now we consider not just threshold exceedance, but the exact categorization. This is shown stratified by the true category of a peak, determined via the empirical distribution of all 396 available peaks. %  (the highest 2.5\% are considered very high, the next 7.5\% high, the next 50\% medium and the remaining 40\% low).
